White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification

### 2015-12-05 papers/wdmerger_I/paper.tex
Fix inconsistent parentheses
Updated description of the source term predictor, and added mention of our new strategy to sum all the source terms into a single array.

Outside of the stars we fill the rest of the domain with a very low density 
ambient gas because our hydrodynamics representation requires the density to be 
non-zero everywhere. This ambient material can create difficulties for the simulation.
In addition to the negative densities or energies at the stellar surfaces mentioned earlier, 
in the rotating reference frame we observe that standing instabilities can create very 
large velocities in the ambient fluid that drag down the global timestep by 
up to an order of magnitude.  To deal with this we employ a ``sponge'' similar 
to that described by \citet{maestro3} for the outer regions of the computational domain. 
After the hydrodynamics update, we apply a damping force to the momentum
equation as follows:
\begin{equation}
  (\rho \mathbf{u})^{n+1} \to \frac{(\rho \mathbf{u}^{n+1})}{1 + (\Delta t / \Delta t_S) f_S},
\end{equation} 
where $\Delta t_S$ is a timescale for the sponge to operate on, and
$f_S$ is the damping factor.  We choose it so that that the sponge is
non-operational inside a radius $r_S$ from the origin, and fully
applied at a radius $r_S^\prime \equiv r_S + \Delta r_S$. We then
smooth the sponge out between $r_S$ and $r_S^\prime$:
\begin{equation}
  f_{S} = \begin{dcases} 0 & r < r_S \\ \frac{1}{2}\left(1 - \text{cos}\left[\pi\left(\dfrac{r - r_S}{\Delta r_S}\right)\right]\right) & r_S \leq r < r_S^\prime \\ 1 & r \geq r_S^\prime. \end{dcases}\label{eq:sponge_frac}
\end{equation}
For the simulations in this paper we set $r_S$ to be 75\% of the 
distance from the origin to the domain boundaries, and $\Delta r_S$ so
that the sponge extends another 10\% of that distance. The resulting profile 
is displayed in \autoref{fig:sponge}. We set $\Delta
t_S = 0.01$ s, which is of the same order as the CFL timestep
for typical problem setups. While the sponge is applied we should be cautious of imputing any physical 
meaning to what is happening in the low-density gas far from the stars.
\begin{figure}
  \centering
  \includegraphics[scale=0.4]{plots/sponge}
  \caption{Radial profile of the hydrodynamical sponge we apply (\autoref{eq:sponge_frac}). 
           We subtract $f_S$ from unity; the value of $1 - f_S$ indicates what happens 
           to the sponged function after the sponge is applied. The sponge has no effect 
           in the inner part of the domain, and is fully applied at the outer edge.\label{fig:sponge}}
\end{figure}

@@

\subsection{Gravity and Rotation Tracing}
\label{sec:gravity_rotation_tracing}

We note a few additional differences between the original PPM
implementation of \citet{ppm} and \castro.  In the original PPM
implementation, the gravitational acceleration was reconstructed as a parabola, and
this was traced under to find the forcing that affects the interface
for each wave.  \castro\ originally followed \citet{ppmunsplit} which instead adds
$(\Delta t/2)g$ to the interface states for velocity at the end of the
reconstruction.  In the current implementation, we do the original
parabolic reconstruction and characteristic tracing of the gravitation
source. This can be controlled in \castro\ with the parameter 
{\tt castro.ppm\_trace\_grav}. Our system with the source appears as:
\begin{equation}
q_t + A(q) q_x = G
\end{equation}
where $G = (0, g, 0)^T$---i.e. the gravitational source only affects
$u$, not $\rho$ or $p$.  Note that in the PPM paper, they put $G$ on 
the left-hand side of the primitive variable equation, so our signs are
opposite.  Our projections are now:
\begin{equation}
\sum_{\nu; \lambda^\enu \ge 0}l^\enu \cdot (\tilde{q} - \mathcal{I}^\enu_+(q) - \tfrac{\Delta t}{2} G) r^\enu
\end{equation}
for the left state, and
\begin{equation}
\sum_{\nu; \lambda^\enu \le 0} l^\enu \cdot (\tilde{q} - \mathcal{I}^\enu_-(q) - \tfrac{\Delta t}{2} G) r^\enu 
\end{equation}
for the right state.  Since $G$ is only non-zero for velocity, only
the velocity changes.  Writing out the sum (and performing the vector products), we
get:
\begin{eqnarray}
u_{i+1/2,L}^{n+1/2} =
   \tilde{u}_+ 
  &-& \frac{1}{2} \left [
      \left (\tilde{u}_+ - \mathcal{I}_+^\evm(u) - \frac{\Delta t}{2} \mathcal{I}^\evm_+(g) \right ) - 
       \frac{\tilde{p}_+ - \mathcal{I}_+^\evm(p)}{C} \right ] \nonumber \\
  &-& \frac{1}{2} \left [
      \left (\tilde{u}_+ - \mathcal{I}_+^\evp(u) - \frac{\Delta t}{2} \mathcal{I}^\evp_+(g) \right ) +
       \frac{\tilde{p}_+ - \mathcal{I}_+^\evp(p)}{C} \right ]
\end{eqnarray}
(The expression in the PPM paper contains $\Delta t G$, not $(\Delta t/2) G$,
but we believe that the factor of $1/2$ is correct.  To see this, notice that if both
waves are moving toward the interface, then the source term that is
added to the interface state is $(\Delta t/4) (\mathcal{I}_+^\evm(g) +
\mathcal{I}_+^\evp(g))$ for the left state, which reduces to $(\Delta
t/2) g$ for constant g---this matches the result from Taylor
expanding to the interface at the half-time (as in \citealt{ppmunsplit}).)

@@

Similarly, we also trace under the centrifugal and Coriolis forces to capture
the effects of rotation on the edge states. For the primitive variables, the rotation
source term is $F_R = (0, f_R, 0)^T$, where $f_R = -2\, {\bm{\omega}} \times \mathbf{u} - {\bm\omega}\times \left({\bm\omega} \times \mathbf{r}\right).$
Just like the gravity source, the rotation source terms only update the velocities. Therefore 
the rotation source terms can be included in a manner identical to that of the gravity source
terms, with the replacement of $G$ with $F_R$ and $g$ with $f_R$ in the above. This is controlled in \castro\
with the parameter {\tt castro.ppm\_trace\_rot}.

\subsection{Source Term Predictor for the Hydrodynamics}

In the original release of \castro\ the value of the gravitational and rotation
source terms used in constructing the edge-states for the hydrodynamics update
was the time-level $n$ value. While this is formally second-order accurate, there
is a better choice one can make. We have information about the trend of these
source terms from previous timesteps, so we can use a predictor method to guess
at a more accurate value of the gravitational and rotational fields at the
$n+1/2$ time-level the hydro is evaluated at. Our method uses a lagged linear
extrapolation. Going into the hydro update, we have both the time-level $n$ and
time-level $n-1$ data for $g$ and $f_R$ (as defined in \autoref{sec:gravity_rotation_tracing}).
From this one can construct a simple linear estimator of (say) the gravitational
acceleration using a backward difference scheme at time-level $n$:
\begin{equation}
  \mathbf{g}(t_n + \Delta t) = \mathbf{g}(t_n) + \Delta t \left.\frac{d\mathbf{g}}{dt}\right|_{n} = \frac{3}{2}\mathbf{g}(t_n) - \frac{1}{2}\mathbf{g}(t_{n-1}).
\end{equation}
Here we have assumed a constant timestep for simplicity, and constructed the time derivative as
$d\mathbf{g} / dt = (\mathbf{g}(t_n) - \mathbf{g}(t_{n-1})) / \Delta t$. We perform this
predictor for the rotation field as well.

>>>>

Outside of the stars we fill the rest of the domain with a very low density 
ambient gas because our hydrodynamics representation requires the density to be 
non-zero everywhere. This ambient material can create difficulties for the simulation.
In addition to the negative densities or energies at the stellar surfaces mentioned earlier, 
in the rotating reference frame we observe that standing instabilities can create very 
large velocities in the ambient fluid that drag down the global timestep by 
up to an order of magnitude.  To deal with this we employ a ``sponge'' similar 
to that described by \citet{maestro3} for the outer regions of the computational domain. 
After the hydrodynamics update, we apply a damping force to the momentum
equation as follows:
\begin{equation}
  (\rho \mathbf{u})^{n+1} \to \frac{(\rho \mathbf{u})^{n+1}}{1 + (\Delta t / \Delta t_S) f_S},
\end{equation} 
where $\Delta t_S$ is a timescale for the sponge to operate on, and
$f_S$ is the damping factor.  We choose it so that that the sponge is
non-operational inside a radius $r_S$ from the origin, and fully
applied at a radius $r_S^\prime \equiv r_S + \Delta r_S$. We then
smooth the sponge out between $r_S$ and $r_S^\prime$:
\begin{equation}
  f_{S} = \begin{dcases} 0 & r < r_S \\ \frac{1}{2}\left(1 - \text{cos}\left[\pi\left(\dfrac{r - r_S}{\Delta r_S}\right)\right]\right) & r_S \leq r < r_S^\prime \\ 1 & r \geq r_S^\prime. \end{dcases}\label{eq:sponge_frac}
\end{equation}
For the simulations in this paper we set $r_S$ to be 75\% of the 
distance from the origin to the domain boundaries, and $\Delta r_S$ so
that the sponge extends another 10\% of that distance. The resulting profile 
is displayed in \autoref{fig:sponge}. We set $\Delta
t_S = 0.01$ s, which is of the same order as the CFL timestep
for typical problem setups. While the sponge is applied we should be cautious of imputing any physical 
meaning to what is happening in the low-density gas far from the stars.
\begin{figure}
  \centering
  \includegraphics[scale=0.4]{plots/sponge}
  \caption{Radial profile of the hydrodynamical sponge we apply (\autoref{eq:sponge_frac}). 
           We subtract $f_S$ from unity; the value of $1 - f_S$ indicates what happens 
           to the sponged function after the sponge is applied. The sponge has no effect 
           in the inner part of the domain, and is fully applied at the outer edge.\label{fig:sponge}}
\end{figure}

@@

\subsection{Source Term Predictor for the Hydrodynamics}
\label{sec:source_term_predictor}

In the original release of \castro\ we used the time-level $n$ value of the 
gravitational and rotation source terms in constructing the edge-states for 
the hydrodynamics update. While this is formally second-order accurate, there
is a better choice one can make. We have information about the trend of these
source terms from previous timesteps, so we can use a predictor method to guess
at a more accurate value of the gravitational and rotational fields at the
$n+1/2$ time-level the hydro is evaluated at. Our method uses a lagged linear
extrapolation. Going into the hydro update, we have both the time-level $n$ and
time-level $n-1$ data for $g$ (as well as the acceleration due to rotation).
From this one can construct a simple linear estimator of (say) the gravitational
acceleration using a backward difference scheme at time-level $n$:
\begin{equation}
  \mathbf{g}^{n+1/2} \approx \mathbf{g}^n + \frac{\Delta t_n}{2} \frac{d\mathbf{g}^n}{dt} \approx \mathbf{g}^n + \frac{\Delta t_n}{2}\left(\frac{\mathbf{g}^n - \mathbf{g}^{n-1}}{\Delta t_{n-1}}\right).
\end{equation}

We note also that we have changed slightly how we use source terms in \castro's 
hydrodynamics. In the original release we would explicitly handle gravity, rotation, 
and user-defined external source terms separately in constructing the edge states. 
At present, we sum all of these source terms prior to starting the hydrodynamics 
update, and use a single source array with data for all components of the state. 
Consequently, we actually do the source-term predictor shown here on this source 
term array, rather than individually on each component of the forcing.

\subsection{Source Term Tracing}
\label{sec:gravity_rotation_tracing}

We note a few additional differences between the original PPM
implementation of \citet{ppm} and \castro.  In the original PPM
implementation, the gravitational acceleration was reconstructed as a parabola, and
this was traced under to find the forcing that affects the interface
for each wave.  \castro\ originally followed \citet{ppmunsplit} which instead adds
$(\Delta t/2)g$ to the interface states for velocity at the end of the
reconstruction.  In the current implementation, we return to the original
parabolic reconstruction and characteristic tracing. In fact, as described 
in \autoref{sec:source_term_predictor}, since we send to the hydro a single 
source term array that holds the sum of all the source terms (including gravity 
and rotation), we do the parabolic reconstruction on the full source term data. 
This can be controlled in \castro\ with the parameter {\tt castro.ppm\_trace\_sources}.

For the following explanation of how the tracing works, we consider only gravity.
Our system with the source appears as:
\begin{equation}
q_t + A(q) q_x = G
\end{equation}
where $G = (0, g, 0)^T$---i.e. the gravitational source only affects
$u$, not $\rho$ or $p$.  Note that in the PPM paper, they put $G$ on 
the left-hand side of the primitive variable equation, so our signs are
opposite.  Our projections are now:
\begin{equation}
\sum_{\nu; \lambda^\enu \ge 0}l^\enu \cdot (\tilde{q} - \mathcal{I}^\enu_+(q) - \tfrac{\Delta t}{2} G) r^\enu
\end{equation}
for the left state, and
\begin{equation}
\sum_{\nu; \lambda^\enu \le 0} l^\enu \cdot (\tilde{q} - \mathcal{I}^\enu_-(q) - \tfrac{\Delta t}{2} G) r^\enu 
\end{equation}
for the right state.  Since $G$ is only non-zero for velocity, only
the velocity changes.  Writing out the sum (and performing the vector products), we
get:
\begin{eqnarray}
u_{i+1/2,L}^{n+1/2} =
   \tilde{u}_+ 
  &-& \frac{1}{2} \left [
      \left (\tilde{u}_+ - \mathcal{I}_+^\evm(u) - \frac{\Delta t}{2} \mathcal{I}^\evm_+(g) \right ) - 
       \frac{\tilde{p}_+ - \mathcal{I}_+^\evm(p)}{C} \right ] \nonumber \\
  &-& \frac{1}{2} \left [
      \left (\tilde{u}_+ - \mathcal{I}_+^\evp(u) - \frac{\Delta t}{2} \mathcal{I}^\evp_+(g) \right ) +
       \frac{\tilde{p}_+ - \mathcal{I}_+^\evp(p)}{C} \right ]
\end{eqnarray}
(The expression in the PPM paper contains $\Delta t G$, not $(\Delta t/2) G$,
but we believe that the factor of $1/2$ is correct.  To see this, notice that if both
waves are moving toward the interface, then the source term that is
added to the interface state is $(\Delta t/4) (\mathcal{I}_+^\evm(g) +
\mathcal{I}_+^\evp(g))$ for the left state, which reduces to $(\Delta
t/2) g$ for constant g---this matches the result from Taylor
expanding to the interface at the half-time (as in \citealt{ppmunsplit}).)

@@
